Iterative Monomorphisation

### 2024-02-14 iterative_monomorphisation.tex
intro

@@

\begin{keywords}
   polymorphism\sep
   higher order logic\sep
   automatic theorem proving
\end{keywords}

@@

%\cite{mono-trans}
%\cite{expr-poly-types}
%\cite{sb-phd}

%what we are trying to do (monomorphisation)\\
This paper presents an iterative method, first introduced in \cite{sb-phd} (section 2.2.1), that allows for higher order polymorphic problems to be approximated monomorphically. This algorithm aims to achieve the following:
   \begin{itemize}
      \item extend higher order theorem provers to polymorphic languages
      \item improve the performance of existing native polymorphic theorem provers
   \end{itemize}

This algorithm is implemented in Zipperpostion, a higher order polymorphic automatic theorem prover written in OCaml. As a result, the E prover, which does not support polymorphism, can be used as a backend to Zipperposition. The viability of this approach is then evaluated, both as an improvement to Zipperposition and as a means to increase the performance of other provers.

\smallskip

[sumarry of results]

\medskip
%basic idea of the algorithm
The algorithm works as follows:
we assume problems to be sets of clauses. All type arguments of the problem are collected and the polymorphic type arguments are matched against the monomorphic ones. This yields new type arguments, both polymorphic and monomorphic. The process is then repeated making use of the newly derived type arguments.

\smallskip

These iterations are required in order to account for inductively defined types. Consider the type constructor \verb|list| which takes a type argument, if we obtain \verb|list(alpha)| where \verb|alpha| is a type variable, it may be possible to generate the types \verb|list(int)|, \verb|list(list(int))|, ... However because new types are obtained through matching, \verb|list(list(int))| can only be obtained once \verb|list(int)| has already been added to the set of monomorphic types. This example also makes it clear that the set of types that can be generated is infinite, this explains why the number of iterations is bounded.

\smallskip

After a set number of iterations, the new type arguments are used to instantiate the type variables in the problem's clauses, generating new monomorphic clauses.

Finally, we ``mangle'' the types in the output clauses using the method presented in \cite{mono-trans}, in order to remove type constructors from the output problem.

This approach sacrifices completeness for the sake of performance. The loss of completeness occurs at the level of the algorithm; a finite number of iterations and at the implementation level; the explosive nature of the procedures requires various bounds which limit the number of type arguments and clauses that are generated.

% target page number: 0.5-1.5
\section{Background}

@@

% target page number: 2-4
% needs to be re-written
\section{High level algorithm}

@@

% target page number: 2-4
\section{Low level algorithm}

@@

% target page number: 2-4
\section{Evaluation}

@@

% target page number: 0.5-1
\section{Related work}

% target page number: 0.5-1
\section{Conclusion}

>>>>

\newcommand\ty[1]{\textsf{#1}}
\newcommand\sym[1]{\textsf{#1}}
\newcommand\var[1]{\mathit{#1}}

@@

\begin{keywords}
   Polymorphism\sep
   monomorphism\sep
   automatic theorem proving
\end{keywords}

@@

{
\obeylines

  * apps are rank-1 polymorphic, but provers are mono
  * there's a gap
  * one solution is to extend the provers to support polymorphism;
    this was done e.g. for Vampire \cite{xxx}
  * but this is a lot of work, and needs to be done for every prover

\bigskip

  * the alternative is to translate polymorphic problems to monomorphic problems
    * two main approaches:
      * complete encoding of polymorphism, using guards or tags,
        as in Blanchette et al.~\cite{xxx} and Bobot et al.~\cite{xxx}
      * incomplete encoding of polymorphism based on iterative monomorphisation
        such as introduced by Böhme \cite[Section 2.2.1]{sb-phd}, whereby type
        variables are heuristically instantiated by concrete types

\bigskip

  * By a type version of Herbrand's theorem, if a set of polymorphic
    formulas is unsatisfiable, then there exists a finite set of monomorphic
    instances of the formulas that is also unsatisfiable. However, this
    finite set cannot be computed \cite{xxx}; hence any approach based on
    instantiation of type variables is incomplete.

\bigskip

  * Böhme's iterative approach is implemented as part of the SMT integration
    \cite{xxx} in Isabelle/HOL. This implementation is also used by
    Sledgehammer \cite{xxx} to interface automatic theorem provers based on
    superposition.

\bigskip

  * The main drawback of Böhme's approach is that it is only briefly documented,
    as a single paragraph in his PhD thesis \cite[Section 2.2.1]{sb-phd}
  * In this paper, we present an algorithm inspired by his description and
    implementation (Section~\ref{sec:high-level-algorithm}).
    * we aimed at mathematical precision, so that people who want to reimplement
      the algorithm can follow this blueprint
    * we also show how to bound the algorithm to avoid explosions in practice
      (Section~\ref{sec:low-level-algorithm}).
      ``the explosive nature of the procedures requires various bounds which
      limit the number of type arguments and clauses that are generated.''

}

The algorithm works as follows. We assume problems to be sets of formulas. All type arguments of the problem are collected, and the polymorphic type arguments are matched against the monomorphic ones. This yields new type arguments, both polymorphic and monomorphic. The process is then repeated using the newly derived type arguments.

These iterations are required to account for $n$-ary type constructors. Consider the unary type constructor \ty{list}. If a formula contains $\ty{list}(\alpha)$, where $\alpha$ is a type variable, it may be possible to generate the types $\ty{list}(\ty{int})$, $\ty{list}(\ty{list}(\ty{int}))$, etc. However, because new types emerge through matching, $\ty{list}(\ty{list}(\ty{int}))$ can be obtained only once $\ty{list}(\ty{int})$ has already been added to the set of monomorphic types. This example also makes it clear that the set of types that can be generated is infinite.

To keep the number of generated formulas finite, we limit the number of iterations. After the iterations are completed, the new type arguments are used to instantiate the type variables in the problem's formulas, generating new monomorphic clauses. Finally, because monomorphic provers generally do not support $n$-ary type constructors, types must be mangled. For example, the compound type $\ty{list}(\ty{int})$ could be mangled to $\ty{list\_int}$.

We implemented this algorithm in Zipperpostion, a higher order automatic theorem prover written in OCaml. Although Zipperposition is polymorphic, it uses the monomorphic prover E as a backend. Our algorithm makes it possible, for the first time, to use E with polymorphic problems. Moreover, our implementation of the algorithm in Zipperposition can be used as a preprocessor to interface with other stand-alone provers.

Our empirical evaluation on TPTP problems \cite{xxx} attempts to answer three questions (Section~\ref{sec:evaluation}):
%
\begin{enumerate}
\item Is the new Zipperposition with the E backend more successful on polymorphic problems than Zipperposition without backend?

\item How competitive are monomorphic provers on monomorphised polymorphic problems?

\item Is iterative monomophisation more effective than the native polymorphism implemented in polymorphic provers?
\end{enumerate}

We find that:
%
\begin{enumerate}
\item Zipperposition benefits substantially from the E backend.

\item E with monomorphisation comes close second to the polymorphic prover Vampire.

\item For Leo-III and Vampire, we find that monomorphisation is indeed more effective than native polymorphism.
\end{enumerate}

% target page number: 0.5-1.5
\section{Background}
\label{sec:background}

@@

% target page number: 2-4
% needs to be re-written
\section{High level algorithm}
\label{sec:high-level-algorithm}

@@

% target page number: 2-4
\section{Low level algorithm}
\label{sec:low-level-algorithm}

@@

% target page number: 2-4
\section{Evaluation}
\label{sec:evaluation}

@@

% target page number: 0.5-1
\section{Related work}
\label{sec:related-work}

% target page number: 0.5-1
\section{Conclusion}
\label{sec:conclusion}
